### 2019-09-20 Manuscript/Manuscript.tex

Indonesia and New Guinea (I\&NG) have an outsized contribution to modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \cite{Gaillardet1999a, Hartmann2009a, Milliman2013a, Hartmann2014a}. There has been a significant increase of subaerially-exposed land area within the region since the Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system \cite{Molnar2015a, Hall2017a, Macdonald2019a}. Concurrently, after the Miocene Climatic Optimum, a cooling trend began ca. 15~Ma and accelerated over the past 4 million years (m.y.) culminating with the onset of Northern Hemisphere glaciation \cite{Shackleton1984a, Zachos2001a}. Many hypotheses have been proposed to explain this cooling trend including changes in ocean/atmosphere circulation (REFS), a decrease in volcanic outgassing (REFS), uplift in the Himalaya (REFS), or an increase in basaltic watersheds in the tropics (REFS). Here we use geologic data to reconstruct the shorelines of I\&NG over the past 20~m.y. (Fig. \ref{fig:shoreline_growth}) and quantify the increase in land area in the tropics. We use these paleoshorelines as boundary conditions within a coupled weathering-climate model to quantify the change in steady-state \pCOtwo associated with this emergence (Fig. \ref{fig:scenario_pCO2}). We find that the decrease in \pCOtwo due to these paleogeographic changes is sufficient to explain long-term climatic cooling over the Neogene. This result highlights the importance of highly weatherable regions in setting Earth's long-term climate state.

Over geologic time-scales, \COtwo enters Earth's ocean/atmosphere system primarily via volcanism and metamorphic degassing, and leaves primarily through the chemical weathering of silicate rocks, which delivers alkalinity and cations to the ocean that sequesters carbon via the precipitation of carbonate sediment \cite{Kump2000a}. Steady-state \pCOtwo is set by the \pCOtwo level at which \COtwo sinks are equal to sources. As \COtwo sinks are removed and \pCOtwo rises, temperature increases and the hydrological cycle is invigorated, causing other weathering sinks to increase until a new steady-state is achieved at higher \pCOtwo. Conversely, the magnitude of the silicate weathering \COtwo sink can be equal to the magnitude of the source at lower \pCOtwo when global weatherability is high \cite{Kump1997a}.

@@

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area in I\&NG and enhanced global weatherability, we use the spatially-resolved GEOCLIM model \cite{Godderis2014a, Godderis2017c}. GEOCLIM consists of a silicate weathering model coupled to a 3D climate model and to a box-model describing the main biogeochemical cycles. GEOCLIM allows the coeval calculations of the CO2 level in the atmosphere and associated climate. Within the silicate weathering component of GEOCLIM, we implement a regolith development model \cite{Gabet2009a} that integrates a climatic dependence on chemical weathering \cite{West2012a}.

@@

The values of four parameters within the silicate weathering component of GEOCLIM \cite{Gabet2009a, West2012a} that modify the dependence of silicate weathering on temperature, runoff, and particle age are poorly constrained (see SI). Rather than prescribing values for these parameters, we instead allow them and the Ca+Mg concentrations of metamorphic and siliciclastic sediment lithologies to vary within reasonable bounds, resulting in 78,000 unique combinations. For each of these combinations, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available\cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating all parameter combinations that result in low $r^{2}$, 3,339 parameter combinations remain. The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of global \COtwo outgassing flux\cite{Gerlach2011a} as they should with the long-term carbon cycle being at steady-state (see SI).

Having calibrated the silicate weathering component of GEOCLIM, we use it to estimate the decrease in steady-state \pCOtwo associated with the emergence of I\&NG. For the climate model component, we use temperature and runoff at three \pCOtwo levels (286, 572, and 1144~ppm) from a subset of the GFDL CM2.0 experiments \cite{Delworth2006a, Delworth2006b} (see SI). The primary strength of these experiments for this analysis is that all non-\COtwo forcings are held constant at values representative of pre-industrial conditions, allowing the effect of \pCOtwo to be isolated. We use terrestrial and marine sedimentary deposits to determine the position of paleoshorelines of I\&NG over the past 20~m.y. and include changes in areas of submerged continental shelves such as the Sunda Shelf that were previously exposed (Fig. \ref{fig:shoreline_growth}; see SI). In New Guinea, emergence in the Middle Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{vanUfford2005a, Cloos2005a, Baldwin2012a}, which drove exhumation of the 1440~km long Irian-Marum-April Ophiolite Belt. Exhumation of the New Guinea Central Range accelerated over the past 4~m.y. associated with slab-breakoff and buoyant uplift \cite{Cloos2005a} while in eastern New Guinea, progressive jamming of the north-dipping subduction zone also caused major uplift over this time interval \cite{vanUfford2005a}. Many of the islands of the Outer Banda Arc emerged after 5~Ma, associated with slab roll-back and collision with the Australian continental margin \cite{Harris2006a, Hall2013b}. These tectonic drivers lead to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C).

To explore the sensitivity of global weatherability and climate state to the emergence of I\&NG, we test three scenarios using GEOCLIM (Fig. \ref{fig:scenario_pCO2}). The ``reduced I\&NG + Sunda Shelf'' scenario represents our best estimate of the subaerial extent of islands in the area at ca. 15~Ma. Given that the Sunda Shelf is presently submerged, across the shelf we distribute felsic and siliciclastic lithologies and assign shallow slope values in a manner that is consistent with existing observations (see SI). While we modify the land area, the position of the tectonic blocks remain fixed since we use a climate model forced with modern geography. While there has been motion of these tectonic blocks since 15~Ma, it has been relatively minimal, and this fixed scenario is a good approximation of the paleogeography (see SI). The ``reduced I\&NG'' scenario is identical to the ``reduced I\&NG + Sunda Shelf'' scenario, but excludes the Sunda Shelf. The ``removed I\&NG'' scenario represents an end-member scenario, in which all islands associated with arc-continent collision in the region are removed.

Using the 3,339 unique parameter combinations, the ``removed I\&NG'' scenario resulted in steady-state \pCOtwo values of 598--729~ppm (5--95 percentile), the ``reduced I\&NG'' scenario resulted in 547--638~ppm \pCOtwo, and the ``reduced I\&NG + Sunda Shelf'' resulted in 533--611~ppm \pCOtwo (Fig. \ref{fig:scenario_pCO2}). These results suggest that without the Neogene increase in subaerially-exposed mountains associated with arc-continent collision, pre-industrial \pCOtwo would have been $\sim$533--611~ppm. These modeled \pCOtwo values resemble the higher end of proxy-based \pCOtwo estimates for the early Miocene \cite{Foster2017a}, indicating that the increase in subaerially-exposed land area and tectonic topography of I\&NG is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values are therefore consistent with Miocene ice sheets on Antarctica followed by Northern Hemisphere ice sheets as \pCOtwo subsequently decreased.

Enhanced weatherability associated with the movement of Deccan Traps flood basalts into the tropical rain belt together with the low-latitude emplacement of the Ethiopian Traps is another hypothesis to explain \COtwo draw-down leading into Oligocene glaciation of Antarctica \cite{Kent2008a, Kent2013a}. To explore the effect of the modern-day surface exposure of basaltic lavas on steady-state \pCOtwo, we replace these mafic volcanics with the Ca+Mg concentration of bulk continental crust in the lithologically-resolved model (Fig. \ref{fig:scenario_pCO2}). The resulting \pCOtwo is \textless500~ppm indicating that, while the presence of mafic rocks in these igneous provinces modulates steady-state \pCOtwo, its contribution is less than that of the wetter and higher-relief I\&NG.

Other hypotheses to explain ice sheet growth over the Neogene invoke changes in ocean/atmosphere circulation including: further climatic isolation of Antarctica due to strengthening of the circumpolar current \cite{Shevenell2004a}; increased atmospheric moisture in the Northern Hemisphere due to intensified thermohaline circulation following Panama Isthmus emergence \cite{Haug1998a}; and cooling of North America resulting from a strengthened Walker Circulation associated with I\&NG emergence \cite{Molnar2015a}. Such changes in ocean/atmosphere circulation are likely to modulate \pCOtwo thresholds for glacial initiation and ice sheet growth \cite{DeConto2008a}. However, the prolonged timescale of the cooling trend since 14~Ma (Fig. \ref{fig:shoreline_growth}C) is most readily attributable to decreasing \pCOtwo associated with evolving geological sources and sinks of carbon, modulated by the silicate weathering feedback \cite{Walker1981a, Raymo1991a, Berner1997a, Kump1997a, Berner2001a}. Uplift in the Himalayas \cite{Raymo1988a} or a decrease in volcanic outgassing \cite{Berner1983a} have also been proposed as drivers for Neogene cooling, although recent analyses show Cenozoic sea floor spreading to be relative uniform \cite{Rowan2016a}. Proponents of the Himalayan hypothesis point toward rising marine \SrSr as being consistent with increasing weathering of the high \SrSr Himalayan lithologies \cite{Raymo1992a}. This rise in \SrSr since ca. 35~Ma is punctuated with a decrease in slope ca. 15~Ma that has been attributed to exhumation of relatively low \SrSr Lesser Himalaya lithologies \cite{Myrow2015a, Colleps2018a}, but could also be partially driven by the emergence of low \SrSr mafic lithologies in I\&NG. While paleogeographic changes elsewhere on Earth would have also modulated geological carbon sources and sinks, the onset of Northern Hemisphere glaciation coincides with increased emergence in I\&NG following 5~Ma. Our coupled weathering-climate model indicates that \pCOtwo change associated with emergence of I\&NG is sufficient to explain climate change over the Neogene, suggesting that the global climate state is particularly sensitive to tropical paleogeography.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/scenario_pCO2.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the six scenarios, each point represents an estimate from one of the 3,339 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched \COtwo consumption estimated from watersheds data (see SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median. The whiskers extend to 1.5$\times$ the interquartile range, or to the minimum/maximum of the estimates if no estimates are below/above this range.}
    \label{fig:scenario_pCO2}
\end{figure}

>>>>

Indonesia and New Guinea (I\&NG) have an outsized contribution to modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \cite{Gaillardet1999a, Hartmann2009a, Milliman2013a, Hartmann2014a}. There has been a significant increase of subaerially-exposed land area within the region since the Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system \cite{Molnar2015a, Hall2017a, Macdonald2019a}. Concurrently, after the Miocene Climatic Optimum, a cooling trend began ca. 15~Ma and accelerated over the past 4 million years (m.y.) culminating with the onset of Northern Hemisphere glaciation \cite{Shackleton1984a, Zachos2001a}. Many hypotheses have been proposed to explain this cooling trend including changes in ocean/atmosphere circulation \cite{Haug1998a, Shevenell2004a, Molnar2015a}, a decrease in volcanic outgassing \cite{Berner1983a}, or uplift in the Himalaya \cite{Raymo1988a}. Here we use geologic data to reconstruct the shorelines of I\&NG over the past 20~m.y. (Fig. \ref{fig:shoreline_growth}) and quantify the increase in land area in the tropics. We use these paleoshorelines as boundary conditions within a coupled weathering-climate model to quantify the change in steady-state \pCOtwo associated with this emergence (Fig. \ref{fig:scenario_pCO2}). We find that the decrease in \pCOtwo due to these paleogeographic changes is sufficient to explain long-term climatic cooling over the Neogene. This result highlights the importance of highly weatherable regions in setting Earth's long-term climate state.

Over geologic time-scales, \COtwo enters Earth's ocean/atmosphere system primarily via volcanism and metamorphic degassing, and leaves primarily through the chemical weathering of silicate rocks, which delivers alkalinity and cations to the ocean that sequesters carbon via the precipitation of carbonate sediment with a smaller sink being organic carbon burial \cite{Kump2000a}. Steady-state \pCOtwo is set by the \pCOtwo level at which \COtwo sinks are equal to sources. As \COtwo sinks are removed and \pCOtwo rises, temperature increases and the hydrological cycle is invigorated, causing other weathering sinks to increase until a new steady-state is achieved at higher \pCOtwo. Conversely, the magnitude of the silicate weathering \COtwo sink can be equal to the magnitude of the source at lower \pCOtwo when global weatherability is high \cite{Kump1997a}.

@@

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area in I\&NG and enhanced global weatherability, we use the spatially-resolved GEOCLIM model \cite{Godderis2014a, Godderis2017c}. GEOCLIM consists of a silicate weathering model coupled to a 3D climate model, allowing for the coeval calculations of \pCOtwo and associated climate. Within the silicate weathering component of GEOCLIM, we implement a regolith development model \cite{Gabet2009a} that integrates a climatic dependence on chemical weathering \cite{West2012a}.

@@

The values of four parameters within the silicate weathering component of GEOCLIM \cite{Gabet2009a, West2012a, Maffre2018a} that modify the dependence of silicate weathering on temperature, runoff, and particle age are poorly constrained (see SI). Rather than prescribing values for these parameters, we instead allow them and the Ca+Mg concentrations of metamorphic and siliciclastic sediment lithologies to vary within reasonable bounds, resulting in 78,000 unique combinations. For each of these combinations, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating all parameter combinations that result in low $r^{2}$, 3,339 parameter combinations remain. The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of global \COtwo outgassing flux \cite{Gerlach2011a} as they should with the long-term carbon cycle being at steady-state (see SI).

Having calibrated the silicate weathering component of GEOCLIM, we use it to estimate the decrease in steady-state \pCOtwo associated with the emergence of I\&NG. For the climate model component, we use temperature and runoff from a subset of the GFDL CM2.0 experiments \cite{Delworth2006a, Delworth2006b} (see SI). The primary strength of these experiments for this analysis is that all non-\COtwo forcings are held constant at values representative of pre-industrial conditions, allowing the effect of \pCOtwo to be isolated. We use terrestrial and marine sedimentary deposits to determine the position of paleoshorelines of I\&NG over the past 15~m.y. and include changes in areas of submerged continental shelves such as the Sunda Shelf that were previously exposed (Fig. \ref{fig:shoreline_growth}; see SI). In New Guinea, emergence in the Middle Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{vanUfford2005a, Cloos2005a, Baldwin2012a}, which drove exhumation of the 1440~km long Irian-Marum-April Ophiolite Belt. Exhumation of the New Guinea Central Range accelerated over the past 4~m.y. associated with slab-breakoff and buoyant uplift \cite{Cloos2005a} while in eastern New Guinea, progressive jamming of the north-dipping subduction zone also caused major uplift over this time interval \cite{vanUfford2005a}. Many of the islands of the Outer Banda Arc emerged after 5~Ma, associated with slab roll-back and collision with the Australian continental margin \cite{Harris2006a, Hall2013b}. These tectonic drivers lead to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C).

To explore the sensitivity of global weatherability and climate state to the emergence of I\&NG, we force GEOCLIM using our best estimates of the subaerial extent of islands in the area at ca. 15, 10, and 5~Ma (``reduced I\&NG + Sunda Shelf'' scenarios; Fig. \ref{fig:scenario_pCO2}). Given that the Sunda Shelf is presently submerged, across the shelf we distribute felsic and siliciclastic lithologies and assign shallow slope values in a manner that is consistent with existing observations (see SI). While we modify the land area, the position of the tectonic blocks remain fixed since we use a climate model forced with modern geography. While there has been motion of these tectonic blocks since 15~Ma, it has been relatively minimal, and this fixed scenario is a good approximation of the paleogeography (see SI). We also test an end member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed I\&NG'' scenario; Fig. \ref{fig:scenario_pCO2}).

Using the 3,339 unique parameter combinations, the ``removed I\&NG'' scenario resulted in steady-state \pCOtwo values of 598--729~ppm (5--95 percentile), and the ``reduced I\&NG + Sunda Shelf'' scenarios resulted in 533--611~ppm, 466--502~ppm, and 394--412~ppm \pCOtwo for 15, 10, and 5~Ma respectively (Fig. \ref{fig:scenario_pCO2}). These results indicate progressive cooling over the Neogene associated with the emergence of I\&NG, and suggest that without it, pre-industrial \pCOtwo would have been $\sim$533--611~ppm. These modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early Miocene ($\sim$200--600~ppm) \cite{Foster2017a}, indicating that the increase in subaerially-exposed land area and tectonic topography of I\&NG is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values are therefore consistent with Miocene ice sheets on Antarctica followed by Northern Hemisphere ice sheets as \pCOtwo subsequently decreased.

The results of our I\&NG scenarios highlight the importance of the combination of topography, runoff, and lithology in setting Earth’s climate state. To independently explore the effect of the modern-day surface exposure of lower-slope basaltic lavas on steady-state \pCOtwo \cite{Kent2008a, Kent2013a}, we replace mafic volcanics associated with the Deccan and Ethiopian Traps with the Ca+Mg concentration of bulk continental crust in the lithologically-resolved model (Fig. \ref{fig:scenario_pCO2}). The resulting \pCOtwo is \textless500~ppm indicating that, while the presence of mafic rocks in these igneous provinces modulates steady-state \pCOtwo, its contribution is less than that of the wetter and higher-relief I\&NG.

Other hypotheses to explain ice sheet growth over the Neogene invoke changes in ocean/atmosphere circulation including: further climatic isolation of Antarctica due to strengthening of the circumpolar current \cite{Shevenell2004a}; increased atmospheric moisture in the Northern Hemisphere due to intensified thermohaline circulation following Panama Isthmus emergence \cite{Haug1998a}; and cooling of North America resulting from a strengthened Walker Circulation associated with I\&NG emergence \cite{Molnar2015a}. Such changes in ocean/atmosphere circulation are likely to modulate \pCOtwo thresholds for glacial initiation and ice sheet growth \cite{DeConto2008a}. However, the prolonged timescale of the cooling trend since 14~Ma (Fig. \ref{fig:shoreline_growth}C) is most readily attributable to decreasing \pCOtwo associated with evolving geological sources and sinks of carbon, modulated by the silicate weathering feedback \cite{Walker1981a, Raymo1991a, Berner1997a, Kump1997a, Berner2001a}.

Uplift in the Himalayas \cite{Raymo1988a} or a decrease in volcanic outgassing \cite{Berner1983a} have also been proposed as drivers for Neogene cooling, although recent analyses show Cenozoic sea floor spreading to be relative uniform \cite{Rowan2016a}. Rising marine \SrSr has been associated with increasing weathering of the high \SrSr Himalayan lithologies \cite{Raymo1992a}. This increase in \SrSr since ca. 35~Ma is followed by a decrease in slope ca. 15~Ma that has been attributed to exhumation of relatively low \SrSr and high \OsOs Lesser Himalaya lithologies \cite{Myrow2015a, Colleps2018a}, or weathering of the Columbia River flood basalts \cite{}, but could also be partially driven by the exhumation of mafic and organic-rich lithologies in I\&NG forearcs during arc-continent collision.

In addition to I\&NG, paleogeographic changes elsewhere on Earth, particularly in the Himalaya and Central America, would have effected geological carbon sources and sinks. Yet, not only does the history of I\&NG emergence coincide with Neogene cooling and the onset of Northern Hemisphere glaciation, but also our coupled weathering-climate model indicates that the associated \pCOtwo change is sufficient to explain the scale of cooling. These results highlight that the Earth's climate state is particularly sensitive to tropical paleogeography.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/scenario_pCO2.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the seven scenarios, each point represents an estimate from one of the 3,339 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched \COtwo consumption estimated from watershed data (see SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to 1.5$\times$ the interquartile range, or to the minimum/maximum of the estimates if no estimates are below/above this range.}
    \label{fig:scenario_pCO2}
\end{figure}
